---
title: "PA1_template.rmd"
output: html_document
---
## Loading and preprocessing the data  
load packages
```{r load, results="hide"}
library(readr)
library(ggplot2)
library(dplyr, warn.conflicts = FALSE)
```
read the data specifying the column types as numeric, Date, numeric
```{r}
data <- read_csv("./activity.csv", col_types = "nDn")
```
## What is mean total number of steps taken per day?  
calculate the total number of steps per day and plot as a histogram  
```{r histogram of total number of steps per day}
stepsPerDay <- summarize(group_by(data, date), total = sum(steps, na.rm=TRUE))
qplot(stepsPerDay$total, data = stepsPerDay, binwidth = 1000, xlab = "Total number of steps per day")
```  

calculated the mean and median number of steps per day and plot  
```{r mean and median number of steps per day}
meanAndMedian <- summarize(group_by(data, date), 
                           mean = mean(steps, na.rm=TRUE), 
                           median = median(steps, na.rm=TRUE))
ggplot(meanAndMedian, aes(date)) + 
    geom_line(aes(y = mean, colour = "mean")) + 
    geom_line(aes(y = median, colour = "median"))
```

Something looks wrong here as the median is 0 for all days and we got warnings for missing values  

## What is the average daily activity pattern?  
Plot the average steps by interval  
```{r Plot the average steps by interval}
meanByInterval <- summarize(group_by(data, interval), 
                           mean = mean(steps, na.rm=TRUE))
ggplot(meanByInterval, aes(interval)) + 
    geom_line(aes(y = mean, colour = "Average steps"))
```  

08:35 has the highest average number of steps  
```{r 08:35 has the highest average number of steps}
meanByInterval[which(meanByInterval$mean == max(meanByInterval$mean)),]
```
## Imputing missing values
there are 2304 missing values in the steps column  
```{r there are 2304 missing values in the steps column}
summary(data)
```  

Now let's try to fill in missing values with the average step for that internval  
```{r fill in missing values with average step for that internval}
inputed.data <- left_join(data, meanByInterval, by = "interval")
inputed.data$steps[is.na(inputed.data$steps)] <- inputed.data$mean[is.na(inputed.data$steps)]
```  

Now plot the histogram of total number of steps taken each day  
```{r histogram of total number of steps taken each day}
stepsPerDay <- summarize(group_by(inputed.data, date), total = sum(steps, na.rm=TRUE))
qplot(stepsPerDay$total, data = stepsPerDay, binwidth = 1000, xlab = "Total number of steps per day")
```  

We see after imputation, that we don't have as many days with 0 steps. Now let's look at the mean and median steps per day with imputed data  
```{r mean and median number of steps per day with imputed data}
meanAndMedian <- summarize(group_by(inputed.data, date), 
                           mean = mean(steps, na.rm=TRUE), 
                           median = median(steps, na.rm=TRUE))
ggplot(meanAndMedian, aes(date)) + 
    geom_line(aes(y = mean, colour = "mean")) + 
    geom_line(aes(y = median, colour = "median"))
```  

After imputation, the graph no longer has breaks, but there are still many days with 0 median  

## Are there differences in activity patterns between weekdays and weekends?  
Let's create a column called group and use that find the average number of steps taken broken out weekday or weekend  
```{r average number of steps taken broken out weekday or weekend}
inputed.data$day <- weekdays(inputed.data$date)
inputed.data$group <- ifelse(inputed.data$day %in% c("Saturday", "Sunday"), "Weekend","Weekday")
meanByInterval <- summarize(group_by(inputed.data, interval, group), 
                           mean = mean(steps, na.rm=TRUE))
qplot(interval, mean, data = meanByInterval, facets = group~., geom = "line")
```
